Generalised gravitational burst generation with Generative Adversarial Networks

### 2021-02-18 Paper_draft/main.tex
addressing p&p reviewer comments

\usepackage[mathlines]{lineno}% Enable numbering of text and display math
\linenumbers\relax % Commence numbering lines

@@

\begin{document}

\title{Generalised gravitational burst generation with Generative Adversarial Networks}

@@

\address{University of Glasgow, Physics \& Astronomy Department, Glasgow G12 8QQ, UK}
%\ead{jordan.mcginn@glasgow.ac.uk}
\vspace{10pt}
%\begin{indented}
%\item[]\commitDATE\\\mbox{\small \commitID}\\\mbox{\dcc}
%\end{indented}

\begin{abstract}
We introduce the use of conditional generative adversarial networks for generalised gravitational wave burst generation in the time domain. Generative adversarial networks are generative machine learning models that produce new data based on the features of the training data set. We condition the network on five classes of time-series signals that are often used to characterise gravitational wave burst searches: sine-Gaussian, ringdown, white noise burst, Gaussian pulse and binary black hole merger. We show that the model can replicate the features of these standard signal classes and, in addition, produce generalised burst signals through interpolation and class mixing. We also present an example application where a convolutional neural network classifier is trained on burst signals generated by our conditional generative adversarial network. We show that a convolutional neural network classifier trained only on the standard five signal classes has a poorer detection efficiency than a convolutional neural network classifier trained on a population of generalised burst signals drawn from the entire signal class space.
\end{abstract}

@@

\begin{comment}
\begin{itemize}
\item Need to introduce GWs - the current state of the field e.g. detections
and LVC papers \ding{51}
\item Introduce burst searches - what's the point of burst searches \ding{51} - lots of references 
\item Discuss the family of burst waveforms currently used and why - not in detail, just
an introduction \ding{51}
\item Introduce ML techniques in GWs \ding{51} - lots of references
\item What this paper does on GANs in 1 paragraph \ding{51}
\item Describe the structure of the paper 
\end{itemize}
\end{comment}

@@

% Discuss the family of burst waveforms currently used and why - not in detail, just
% an introduction 
%
GW burst detection algorithms~\cite{drago2020coherent,Klimenko_2008, Aso_2008} are tested
and tuned using modelled waveforms that have easy to define parameters and share characteristics of real bursts that aim to simulate a GW passing between
detectors. Such waveforms include sine-Gaussians: a
Gaussian modulated sine wave that is characterised by its central frequency and
decay parameter. Bandlimited white noise bursts: white noise that is contained
within a certain frequency range. Ringdowns: which mimic the damped
oscillations after a \ac{CBC} merger. A Gaussian pulse: a short exponential increase then decrease in amplitude and a binary black hole inspiral.
% cut this text if not going to be used
%
\begin{comment}Such waveforms may have
long-duration, short bandwidth (ringdowns), long-duration, large bandwidth
(inspirals) and many algorithms make use of sine-Gaussians: a Gaussian
modulated sine wave that is characterised by it's central frequency and narrow
bandwidth.~\chris{not the time to try to describe the types of burst waveforms.
Also, be careful with satying things like ringdowns have narrow bandwidth. I
know we specify a single frequency but becuase of the short duration, the
signal is broad band. Look at the FFT} This makes it a great tool for
diagnosing LIGOs sensitivity to frequency.~\chris{strange unfinished sentence.} 
\end{comment}

% Introduce ML techniques in GWs
%
With the expectation that there will be many more GW detections in the
future, there is a growing need for fast and efficient GW analysis methods
to match the rising number of detections. While still in its infancy, the application of \ac{ML} to GW analyses has already shown great potential in areas of detection~\cite{Gabbard2017,Gebhard_2019,Krastev_2020}, where these techniques have matched the sensitivity of matched filtering for Advanced LIGO and Advanced Virgo GW searches. Similarly, for unmodelled burst search the flexibility of \ac{ML} algorithms has been shown to be a natural and sensitive approach to detection~\cite{2020arXiv200914611S}. Progress has also been made in identifying and classifying detector noise transients or  ``glitches''~\cite{Bahaadini, George_2018,Razzano_2018, 2020arXiv200801262G} and in Bayesian parameter estimation~\cite{gabbard2019bayesian,Chua:2019,green2020gravitationalwave} where \ac{ML} techniques can recover parameters of a GW signal significantly faster than standard methods. Long duration signals like continuous GW require long observing times and therefore have large amounts of data needing to be processed. Current \ac{ML} approaches~\cite{2020PhRvD.102b2005D, 2019PhRvD.100d4009D, 2020arXiv200708207B} are particularly well suited to dealing with this as once trained the searches can be performed quickly.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Machine learning} \label{ML overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{comment}
\begin{itemize}
\item Describe GANs in detail but really focus on the fact that the reader is a
GW data analyst - not a computer scientist \ding{51}
\item A diagram would be very useful \ding{51}
\item Do not discuss our specific case here - just stay general \ding{51}
\item A subsection on the specific advanced flavour of GAN that you are using
here - motivate this choice. \ding{51}
\end{itemize}
\end{comment}

% introduce GANs
%
\subsection{Artificial neural networks}

@@

\caption{Neural Networks (a) A single neuron taking a vector of inputs and
returning a single output based on the weights, bias, and activation function
of the network. (b) A selection of activation functions used in this study. The sigmoid (solid purple), rectified linear unit (ReLU)~\cite{relu} (dashed red) and leaky rectified linear unit~\cite{Maas2013RectifierNI} (dotted blue). (c) A an example of a neural network containing two hidden layers that performs a mapping of an input vector to a single output.}
\end{figure}
%
% Introduce basic neural networks - a perceptron layer
%
\begin{comment}
Neural networks are the quintessential~\chris{really? quintessential?} \jordan{It's a perfectly cromulent word https://www.imdb.com/title/tt0701155/} \ac{ML}
algorithm that aims to approximate a function. 
\end{comment}

\ac{ML} algorithms aim to learn apparent relationships held within given data or `training
data' in order to make accurate predictions without the need for additional
programming. A common approach in \ac{ML} relies on the model learning from
past experience to make decisions on future events. Artificial neural networks are universal function approximators that are built from many single
processing units called neurons. The simplest neural network is the perceptron
layer~\cref{fig:perceptron} which shows a single neuron that takes a vector of real
inputs $x_{i},\ldots, x_{n}$
and maps them to an output according to the linear function, 
%
\begin{align}
f(x) = \sigma\left(\sum_i w_i x_i + b\right),
\label{eqn:neuron}
\end{align}
%
where $w$ and $b$ are the weights and bias and $\sigma$ denotes the activation function. The weights are numbers which can be thought of as the strength between connected neurons. The output of a neuron is defined by its activation function which controls how the neuron `fires' depending on its input. Some examples of commonly used activation functions are shown in~\cref{fig:activation}. It is often useful to introduce a bias, $b$, such that the neuron remains inactive above zero but is active when the sum reaches a defined threshold.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Generative Adversarial Networks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% basic intro to GANs
%
A subset of deep learning that has seen fruitful development in recent years
are generative adversarial networks \acp{GAN}~\cite{Goodfellow2014}. These unsupervised algorithms learn patterns in a
given training dataset using an adversarial process. The generations from
\acp{GAN} are currently state-of-the-art in fields such as high quality image
fidelity~\cite{brock2018large,karras2019analyzing}, text-to-image
translation~\cite{reed2016generative}, and video
prediction~\cite{liang2017dual} as well as time-series
generations~\cite{esteban2017realvalued}.
%
% Basic components of a GAN
%
\acp{GAN} train two competing neural networks, consisting of a discriminator
network that is set up to distinguish between real and fake data and a
generator network that produces fake versions of the real data. The generator model performs a mapping from a fixed length vector $\mathbf{z}$ to its
representation of the data. The input vector is drawn randomly from a Gaussian distribution which is referred to as a latent space comprised of latent variables. The latent space is a compressed representation of a data distribution which the generator applies meaning to during training. Sampling points from this space allows the generator to produce a variety of different generations, with different points corresponding to different features in the generations. The discriminator maps its input $\mathbf{x}$ to a probability that the input came from either the training (real) data or
generator (fake).
\begin{comment}
Chris' definition of latent space if needed: It maps the intrinsic
variation of the training space onto the Gaussian latent space distribution.
\end{comment}
%
% training a GAN
%
During training, the discriminator and generator are updated using batches of data. Random latent vectors are given to the generator to produce a batch of fake samples and an equal batch of real samples is taken from the training data. The discriminator makes predictions on the real and fake samples and the model is updated through minimising the binary cross-entropy function \cite{Goodfellow-et-al-2016}
%
\begin{equation}
    L = -y \log(\hat{y}) - (1 - y) \log(1-\hat{y}),
    \label{eqn:crossentropy}
\end{equation}
where $\hat{y}$ is the network prediction and $y$ is the true output. While training the discriminator, $D$, on real data, we set $y = 1$ and $\hat{y} = D(\mathbf{x})$ which from \cref{eqn:crossentropy} gives $L(D(\mathbf{x}),1) = \log(D(\mathbf{x}))$. While training on fake data produced by the generator, $G$, $y = 0$ and $\hat{y} = D(G(\mathbf{z}))$ and so, $L(D(G(\mathbf{z})),0) = \log(1-(D(G(\mathbf{z}))))$. Since the objective of the discriminator is to correctly classify fake and real data these equations should be maximised, while the goal of the generator should be to minimize these equations. This gives the \ac{GAN} loss as
%
\begin{equation}
   \mathop{\text{min}}_{G}  \mathop{\text{max}}_{D} V(D,G) = \mathbb{E}_{\mathbf{x} \sim p_{\text{\text{data}}}(\mathbf{x})} [\text{log} D(\mathbf{x})] + \mathbb{E}_{\mathbf{z} \sim p_{\mathbf{z}}(\mathbf{z})} [\text{log}(1-D(G(\mathbf{z})))],
 \label{equation:GANloss}
 \end{equation}
where $p_{\text{\text{data}}}(\mathbf{x})$ is the distribution of real data and $p_{\text{z}}(\mathbf{z})$ is the latent distribution.

\subsection{Training stages}
%
Training a GAN is involves updating both the discriminator and generator in stages. First, the discriminator is updated using real instances from the training set. We set the true label $y=1$ and calculate the loss with respect to the predictions $\hat{y}$ via \cref{equation:GANloss}. Stochastic gradient descent is used to minimize the loss which has reduced to $L_D(\textrm{real}) = -\log(\hat{y})$. The discriminator is then trained on fake instances taken from the generator where we set $y=0$ and minimize $L_D(\textrm{fake}) = -\log(1-\hat{y})$. To train the generator, we use a composite model of the generator and discriminator and allow the gradients to flow through this entire model. At this step we set the labels for the generator as real $y=1$ and minimize $L_G = -\log(\hat{y})$. The act of switching the labels and minimizing this form of loss is identical to the previously mentioned generator loss but is beneficial as in practice this function provides more useful gradients. This tweak to the generator loss is called non-saturating generator loss and was reported in the original \ac{GAN} paper \cite{Goodfellow2014}. It was also shown in that paper that if the generator and discriminator can no longer improve, then the discriminator can no longer distinguish between real and fake i.e. $D(x) = \frac{1}{2}$.

@@

\begin{figure}[h!]
    \begin{subfigure}{.5\textwidth}
     \centering
        \input{figures/GAN_diagram_vertical}
        \caption{GAN}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
     \centering
        \input{figures/ACGAN_diagram_vertical}
        \caption{CGAN}
    \end{subfigure}
    \caption{Comparison of the original GAN method and the
conditional-GAN method. $\textbf{G}$ and $\textbf{D}$ denote the generator and discriminator neural networks respectively while $\textbf{X}~\text{real}$ and $\textbf{X}~\text{fake}$ represent samples drawn from the training set and the generated set. For CGANs the training data requires a label denoting
its class that is also fed to the generator which then learns to generate
waveforms based on the input label.}
\begin{comment}
~\chris{Overall it's a bit basic and doesn't indicate either the
2 competing training steps or any aspect of the loss or specify that $x$ and
$z$ (and $c$) are drawn from distributions already defined in the text. Maybe it's OK without
that. As for the CGAN, I'm troubled by the fact that there is only one $c$ box.
It implies that the same $c$ value is given to each real and fake $x$ data when
in fact each $x$ sample has it's own randomly drawn $c$ value.}
\end{comment}
\label{fig:gan_comparison}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Training data and architecture} \label{Method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{comment}
\begin{itemize}
\item Need to introduce the scheme you propose to use
\item A paragraph or subsection on the data generation being very clear on all
5 waveform models and the prior parameter space for each \ding{51}
\item A subsection on the design of the network architecture \ding{51}
\item A subsection on the "box" and why we implement it \ding{51}
\item A subsection on the training of the network - give rough timings and rule
of thumb decisions made
\item Do not discuss the results here 
\end{itemize}
\end{comment}

% introduce the training data
%
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/training-sample.pdf}
    \caption{Examples of the five different waveforms that were used in training the \ac{GAN} for this study. Values of the parameters were selected randomly from uniform distributions from \cref{Tab:training_parms}.}
    \label{fig:training_waveforms}
\end{figure}
% introduce the signal models
%
We propose a signal generation scheme using a \ac{CGAN} trained on burst-like waveforms. We call this \texttt{McGANn}\footnote{https://github.com/jmcginn/McGANn} and it is a \ac{CGAN} trained on five signal classes each spanning a range of prior signal parameters. The signal classes are:

% list the 5 waveform classes
%
\begin{itemize}
%
\item {\bf Sine-Gaussian}: $h_{\text{SG}}(t) = A \exp\left[ - (t-t_{0})^2 /
\tau^2 \right] \sin (2 \pi f_0 (t-t_0) + \phi)$, a sine wave with a Gaussian envelope characterised by a central frequency $f_0$, amplitude $A$, time of arrival $t_{0}$ and phase $\phi$ which is uniformly sampled between $[0, 2\pi]$.
%
\item {\bf Ringdown}: $h_{\text{RD}}(t) = A \exp \left[-{(t-t_0)} / {\tau}
\right] \sin(2 \pi f_0 (t-t_0) + \phi)$, with frequency $f_0$ and duration $\tau$, amplitude $A$, time of arrival $t_{0}$ and phase $\phi$ which is uniformly sampled between $[0, 2\pi]$. 
%
\item {\bf White noise bursts}: $h_{\text{WN}}(t_j) = Ag_j\exp\left[ -
(t-t_{0})^2 / \tau^2 \right]$ where $g_j$ are drawn from a zero mean unit
variance Gaussian distribution with a Gaussian envelope of duration $\tau$.
%
\item {\bf Gaussian pulse}: $h_{\text{GP}}(t) = \exp(-t^2 / \tau^2)$ with
duration parameter $\tau$.
%
\item {\bf Binary black hole}: Simulated using the IMRPhenomD
waveform~\cite{Khan_2016} routine from LALSuite~\cite{lalsuite} which models
the inspiral, merger and ringdown of a \ac{BBH} waveform. The component masses
lie in the range of [30,70] $\textrm{M}_{\odot}$ with zero spins and we fix
$m_1>m_2$. The mass distribution is approximated by a power law with
index of 1.6~\cite{Abbott_2019}. The inclinations are drawn
such that the cosine of the angles lies uniformly in the range [-1,1] and we only use the plus polarisation.
%
\end{itemize}
%
The location of the peak amplitude of the waveforms (corresponding to the
mid-points of all but the ringdown and \ac{BBH} classes) are randomly drawn from a uniform distribution to
be within $[0.4, 0.6]$ seconds from the start of the 1 second time interval and all
training waveforms are sampled at 1024 Hz.  The parameter prior ranges are
defined in~\cref{Tab:training_parms} and a sample of training waveforms are shown in \cref{fig:training_waveforms}. All training data is rescaled such that their amplitudes peak at 1.

\begin{table}[!h]
\centering
\caption{The parameters used in generating the training data. Each parameter is drawn uniformly in the below ranges.}
%\footnotesize
\begin{tabular}{@{} l l l l l l }
\br
\hline
 Waveform & Central frequency  & Decay & Central time epoch & Mass range \\
 & (Hz) & (s) & (s) & ($\textrm{M}_{\odot}$) \\
\mr
Sine-Gaussian & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & -  \\  
Ringdown & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & - \\
white noise burst & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & -  \\
Gaussian pulse & - & 0.004 - 0.03 & 0.4 - 0.6 & -  \\
BBH & - & - & - & 30 - 70  \\
 \br
\end{tabular}\\
\label{Tab:training_parms}
\end{table}
\normalsize

@@

\begin{comment}
~\chris{Also, discuss the corresponding figure! There has to be comment on the
features of simplex vs box generations as well as general comments about how
the GAN likes to mix signals up. Pick on specific examples and comment on
interesting or annoying features.We can expect a priori that the uniformly sampled space (which contains the simplex as a subspace) will allow for some level of extrapolation.}

\michael{You could also point to \cref{fig:gen_signals} since you have figures for Uniform and Simplex}

~\chris{I have an idea for the plots. Could we have a colour coding system
where each class gets a colour and then when you mix them you plot it with the
mixed colours - you average the RGB values. Maybe everything would look brown?
Worth a try?}
\end{comment}

@@

\begin{comment}
\begin{itemize}
\item Summarise the paper
\item Dedicate a paragraph to each of the key results discussed in the previous
section
\item Have at least one paragraph on the future directions of this work
\item Conclude with a positive paragrpah about the potential uses and impact of
the approach.
\end{itemize}
\end{comment}

>>>>

\usepackage[mathlines]{lineno}% Enable numbering of text and display math
%\linenumbers\relax % Commence numbering lines

@@

\begin{document}
\title{Generalised gravitational burst generation with Generative Adversarial Networks}

@@

\address{SUPA, School of Physics and Astronomy, University of Glasgow, Glasgow G12 8QQ, United Kingdom}
%\ead{jordan.mcginn@glasgow.ac.uk}
\vspace{10pt}
%\begin{indented}
%\item[]\commitDATE\\\mbox{\small \commitID}\\\mbox{\dcc}
%\end{indented}
\begin{abstract}
We introduce the use of conditional generative adversarial networks for generalised gravitational wave burst generation in the time domain. Generative adversarial networks are generative machine learning models that produce new data based on the features of the training data set. We condition the network on five classes of time-series signals that are often used to characterise gravitational wave burst searches: sine-Gaussian, ringdown, white noise burst, Gaussian pulse and binary black hole merger. We show that the model can replicate the features of these standard signal classes and, in addition, produce generalised burst signals through interpolation and class mixing. We also present an example application where a convolutional neural network classifier is trained on burst signals generated by our conditional generative adversarial network. We show that a convolutional neural network classifier trained only on the standard five signal classes has a poorer detection efficiency than a convolutional neural network classifier trained on a population of generalised burst signals drawn from the combined signal class space.
\end{abstract}

@@

@@

% Discuss the family of burst waveforms currently used and why - not in detail, just
% an introduction 
%
GW burst detection algorithms~\cite{drago2020coherent,Klimenko_2008, Aso_2008} are tested
and tuned using modelled waveforms that have easy to define parameters and share characteristics of real bursts that aim to simulate a GW passing between
detectors. Such waveforms include sine-Gaussians: a
Gaussian modulated sine wave that is characterised by its central frequency and
decay parameter. Bandlimited white noise bursts: white noise that is contained
within a certain frequency range. Ringdowns: which mimic the damped
oscillations after a \ac{CBC} merger. A Gaussian pulse: a short exponential increase then decrease in amplitude and a binary black hole inspiral.
% cut this text if not going to be used
%
% Introduce ML techniques in GWs
%
With the expectation that there will be many more GW detections in the
future, there is a growing need for fast and efficient GW analysis methods
to match the rising number of detections. While still in its infancy, the application of \ac{ML} to GW analyses has already shown great potential in areas of detection~\cite{Gabbard2017,Gebhard_2019,Krastev_2020}, where these techniques have matched the sensitivity of matched filtering for Advanced LIGO and Advanced Virgo GW searches. Similarly, for unmodelled burst search the flexibility of \ac{ML} algorithms has been shown to be a natural and sensitive approach to detection~\cite{2020arXiv200914611S}. Progress has also been made in identifying and classifying detector noise transients or  ``glitches''~\cite{Bahaadini, George_2018,Razzano_2018, 2020arXiv200801262G} and in Bayesian parameter estimation~\cite{gabbard2019bayesian,Chua:2019,green2020gravitationalwave} where \ac{ML} techniques can recover parameters of a GW signal significantly faster than standard methods. Long duration signals like continuous GW require long observing times and therefore have large amounts of data needing to be processed. Current \ac{ML} approaches~\cite{2020PhRvD.102b2005D, 2019PhRvD.100d4009D, 2020arXiv200708207B} are particularly well suited to dealing with this as once trained the searches can be performed quickly.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Machine learning} \label{ML overview}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% introduce GANs
%
\subsection{Artificial neural networks}

@@

\caption{Neural Networks (a) A single neuron taking a vector of inputs and
returning a single output based on the weights, bias, and activation function
of the network. (b) A selection of activation functions used in this study. The sigmoid (solid purple), rectified linear unit (ReLU)~\cite{relu} (dashed red) and leaky rectified linear unit~\cite{Maas2013RectifierNI} (dotted blue). (c) A an example of a neural network containing two hidden layers that performs a mapping of an input vector to a single output.}
\end{figure}
%
% Introduce basic neural networks - a perceptron layer
%
\ac{ML} algorithms aim to learn apparent relationships held within given data or `training
data' in order to make accurate predictions without the need for additional
programming. A common approach in \ac{ML} relies on the model learning from
past experience to make decisions on future events. Artificial neural networks are universal function approximators that are built from many single
processing units called neurons. The simplest neural network is the perceptron
layer~\cref{fig:perceptron} which shows a single neuron that takes a vector of real
inputs $x_{i},\ldots, x_{n}$
and maps them to an output according to the linear function, 
%
\begin{align}
f(x) = \sigma\left(\sum_i w_i x_i + b\right),
\label{eqn:neuron}
\end{align}
%
where $w$ and $b$ are the weights and bias and $\sigma$ denotes the activation function. The weights are numbers which can be thought of as the strength between connected neurons. The output of a neuron is defined by its activation function which controls how the neuron `fires' depending on its input. Some examples of commonly used activation functions are shown in~\cref{fig:activation}. It is often useful to introduce a bias, $b$, such that the neuron remains inactive above zero but is active when the sum reaches a defined threshold.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Generative Adversarial Networks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% basic intro to GANs
%
A subset of deep learning that has seen fruitful development in recent years
are generative adversarial networks \acp{GAN}~\cite{Goodfellow2014}. These unsupervised algorithms learn patterns in a
given training dataset using an adversarial process. The generations from
\acp{GAN} are currently state-of-the-art in fields such as high quality image
fidelity~\cite{brock2018large,karras2019analyzing}, text-to-image
translation~\cite{reed2016generative}, and video
prediction~\cite{liang2017dual} as well as time-series
generations~\cite{esteban2017realvalued}.
%
% Basic components of a GAN
%
\acp{GAN} train two competing neural networks, consisting of a discriminator
network that is set up to distinguish between real and fake data and a
generator network that produces fake versions of the real data. The generator model performs a mapping from a fixed length vector $\mathbf{z}$ to its
representation of the data. The input vector is drawn randomly from a Gaussian distribution which is referred to as a latent space comprised of latent variables. The latent space is a compressed representation of a data distribution which the generator applies meaning to during training. Sampling points from this space allows the generator to produce a variety of different generations, with different points corresponding to different features in the generations. The discriminator maps its input $\mathbf{x}$ to a probability that the input came from either the training (real) data or
generator (fake).
%
% training a GAN
%
During training, the discriminator and generator are updated using batches of data. Random latent vectors are given to the generator to produce a batch of fake samples and an equal batch of real samples is taken from the training data. The discriminator makes predictions on the real and fake samples and the model is updated through minimising the binary cross-entropy function \cite{Goodfellow-et-al-2016}
%
\begin{equation}
    L = y \log(\hat{y}) - (1 - y) \log(1-\hat{y}),
    \label{eqn:crossentropy}
\end{equation}
where $\hat{y}$ is the network prediction and $y$ is the true output. While training the discriminator, $D$, on real data, we set $y = 1$ and $\hat{y} = D(\mathbf{x})$ which from \cref{eqn:crossentropy} gives $L(D(\mathbf{x}),1) = \log(D(\mathbf{x}))$. While training on fake data produced by the generator, $G$, $y = 0$ and $\hat{y} = D(G(\mathbf{z}))$ and so, $L(D(G(\mathbf{z})),0) = \log(1-(D(G(\mathbf{z}))))$. Since the objective of the discriminator is to correctly classify fake and real data these equations should be maximised, while the goal of the generator should be to minimize these equations. This gives what is know as the \ac{GAN} value function as
%
\begin{equation}
   \mathop{\text{min}}_{G}  \mathop{\text{max}}_{D} V(D,G) = \mathbb{E}_{\mathbf{x} \sim p_{\text{\text{data}}}(\mathbf{x})} [\text{log} D(\mathbf{x})] + \mathbb{E}_{\mathbf{z} \sim p_{\mathbf{z}}(\mathbf{z})} [\text{log}(1-D(G(\mathbf{z})))],
 \label{equation:GANloss}
 \end{equation}
where $p_{\text{\text{data}}}(\mathbf{x})$ is the distribution of real data and $p_{\text{z}}(\mathbf{z})$ is the latent distribution.

\subsection{Training stages}
%
Training a GAN involves updating both the discriminator and generator in stages. First, the discriminator is updated using real instances from the training set. We set the true label $y=1$ and calculate the loss with respect to the predictions $\hat{y}$ via \cref{eqn:crossentropy}. Stochastic gradient descent is used to maximise the loss which has reduced to $L_D(\textrm{real}) = \log(\hat{y})$. The discriminator is then trained on fake instances taken from the generator where we set $y=0$ and maximise $L_D(\textrm{fake}) = \log(1-\hat{y})$. To train the generator, we use a composite model of the generator and discriminator and allow the gradients to flow through this entire model. Following on from what was described before, to train the generator we set y = 0 and minimise $L_{G}$(fake) = $\log(1-\hat{y})$. During early stages of training the generator produces poor generations and so D can easily determine them as fake i.e. $\hat{y})$~0. This leads $L_{G}$ to tend to 0 and we encounter the \textit{vanishing gradient problem}, where the gradients become so small that the weights can no longer be updated. A solution to this problem involves changing the generator loss to maximise $L_{G}$(fake) = $\log(\hat{y})$ or equivalently continue to minimise $L_{G}$(fake) = $\log(1-\hat{y})$ and simply switch the $y$ label to 1. This tweak to the generator loss is called non-saturating generator loss and was reported in the original \ac{GAN} paper \cite{Goodfellow2014}. It was also shown in that paper that if the generator and discriminator can no longer improve, then the discriminator can no longer distinguish between real and fake i.e. $D(x) = \frac{1}{2}$.

@@

\begin{figure}[h!]
    \begin{subfigure}{.5\textwidth}
     \centering
        \input{figures/GAN_diagram_vertical}
        \caption{GAN}
    \end{subfigure}
    \begin{subfigure}{.5\textwidth}
     \centering
        \input{figures/ACGAN_diagram_vertical}
        \caption{CGAN}
    \end{subfigure}
    \caption{Comparison of the original GAN method and the
conditional-GAN method. $\textbf{G}$ and $\textbf{D}$ denote the generator and discriminator neural networks respectively while $\textbf{X}~\text{real}$ and $\textbf{X}~\text{fake}$ represent samples drawn from the training set and the generated set. For CGANs the training data requires a label denoting
its class that is also fed to the generator which then learns to generate
waveforms based on the input label.}
\label{fig:gan_comparison}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Training data and architecture} \label{Method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% introduce the training data
%
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/training-sample.pdf}
    \caption{Examples of the five different waveforms that were used in training the \ac{GAN} for this study. Values of the parameters were selected randomly from uniform distributions from \cref{Tab:training_parms}.}
    \label{fig:training_waveforms}
\end{figure}
% introduce the signal models
%
We propose a signal generation scheme using a CGAN trained on burst-like waveforms which we call \texttt{McGANn}\footnote{https://github.com/jmcginn/McGANn}. \texttt{McGANn} is trained on five signal classes which are used to characterise the sensitivity of gravitational wave burst searches (see for example \cite{PhysRevD.100.024017}).

% list the 5 waveform classes
%
\begin{itemize}
%
\item {\bf Sine-Gaussian}: $h_{\text{SG}}(t) = A \exp\left[ - (t-t_{0})^2 /
\tau^2 \right] \sin (2 \pi f_0 (t-t_0) + \phi)$, a sine wave with a Gaussian envelope characterised by a central frequency $f_0$, amplitude $A$, time of arrival $t_{0}$ and phase $\phi$ which is uniformly sampled between $[0, 2\pi]$.
%
\item {\bf Ringdown}: $h_{\text{RD}}(t) = A \exp \left[-{(t-t_0)} / {\tau}
\right] \sin(2 \pi f_0 (t-t_0) + \phi)$, with frequency $f_0$ and duration $\tau$, amplitude $A$, time of arrival $t_{0}$ and phase $\phi$ which is uniformly sampled between $[0, 2\pi]$. 
%
\item {\bf White noise bursts}: $h_{\text{WN}}(t_j) = Ag_j\exp\left[ -
(t-t_{0})^2 / \tau^2 \right]$ where $g_j$ are drawn from a zero mean unit
variance Gaussian distribution with a Gaussian envelope of duration $\tau$.
%
\item {\bf Gaussian pulse}: $h_{\text{GP}}(t) = \exp(-t^2 / \tau^2)$ with
duration parameter $\tau$.
%
\item {\bf Binary black hole}: Simulated using the IMRPhenomD
waveform~\cite{Khan_2016} routine from LALSuite~\cite{lalsuite} which models
the inspiral, merger and ringdown of a \ac{BBH} waveform. The component masses
lie in the range of [30,70] $\textrm{M}_{\odot}$ with zero spins and we fix
$m_1>m_2$. The mass distribution is approximated by a power law with
index of 1.6~\cite{Abbott_2019}. The inclinations are drawn
such that the cosine of the angles lies uniformly in the range [-1,1] and we only use the plus polarisation.
%
\end{itemize}
%
The location of the peak amplitude of the waveforms (corresponding to the
mid-points of all but the ringdown and \ac{BBH} classes) are randomly drawn from a uniform distribution to
be within $[0.4, 0.6]$ seconds from the start of the 1 second time interval and all
training waveforms are sampled at 1024 Hz.  The parameter prior ranges are
defined in~\cref{Tab:training_parms} and a sample of training waveforms are shown in \cref{fig:training_waveforms}. All training data is rescaled such that their amplitudes peak at 1.
\begin{table}[!h]
\centering
\caption{The parameters used in generating the training data. Each parameter is drawn uniformly in the below ranges.}
%\footnotesize
\begin{tabular}{@{} l l l l l l }
\br
\hline
 Waveform & Central frequency  & Decay & Central time epoch & Mass range \\
 & (Hz) & (s) & (s) & ($\textrm{M}_{\odot}$) \\
\mr
Sine-Gaussian & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & -  \\  
Ringdown & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & - \\
white noise burst & 70 - 250 & 0.004 - 0.03 & 0.4 - 0.6 & -  \\
Gaussian pulse & - & 0.004 - 0.03 & 0.4 - 0.6 & -  \\
BBH & - & - & - & 30 - 70  \\
 \br
\end{tabular}\\
\label{Tab:training_parms}
\end{table}
\normalsize

With the exception of the binary black hole waveforms, the signal classes described above are analytic proxy waveforms to gravitational wave signals expected from various burst gravitational wave sources. For example, numerical relativity simulations show that rapidly rotating stellar core collapse emit gravitational waves that look like sine gaussians with small time constants (low Q). Additionally, gravitational waves from hyperbolic black hole encounters look very similar to sine gaussians and gaussian pulses. Ringdown signals can be emitted by excited isolated neutron stars, for example, after a pulsar glitch and white noise burst signals mimic the stochastic nature of gravitational wave signals emitted by neutrino driven stellar core collapse.

@@

@@
